Indirect evidence white is actually a funny color

### 2019-03-25 article.tex
To 80 chars
EOD

% RJCB: I think this title is better: 
% (1) tree prior importantance is the goal, the error is just the means
% (2) no need to specify BEAST2, it's also just a means
\title{
  pirouette:
  Quantifying the importance of a tree prior in Bayesian phylogenetics
}
% \title{pirouette: Determining the error BEAST2 makes in inferring a phylogeny}

\author[1]{Rich\`el J.C. Bilderbeek}
\author[1]{Giovanni Laudanno}
\author[1]{Rampal S. Etienne}
\affil[1]{Groningen Institute for Evolutionary Life Sciences, University of 
Groningen, Groningen, The Netherlands}

@@

\begin{abstract}

  \textbf{1. }
    BEAST2 is a popular Bayesian phylogenetics software tool,
    that takes an alignment of characters (usually nucleotides) 
    and an inference model to create a
    posterior of jointly-estimated phylogenies and model parameter 
    estimates. An important ingredient of BEAST2 is 
    the species tree prior, 
    due to which the tool can be used for inference of 
    macroevolutionary processes underlying the species tree. 
    However, only models that allow for fast computation of 
    the prior probability are possible in practice. 
    This begs the question of how accurate the tree estimation is 
    when the real macroevolutionary processes are different 
    from those assumed in BEAST2. \\
  \textbf{2. }
    Here we present \verb;pirouette;, 
    a free, libre and open-source R package that assesses 
    the inference error BEAST2 makes based on a known/true 
    phylogeny, for example generated by a new 
    macroevolutionary diversification model. \\
  \textbf{3. }
    We describe \verb;pirouette;'s usage and the biological scientific
    question it can answer, including full examples. \\
  \textbf{4. }
    Last, we discuss the results obtained by the examples. \\
\end{abstract}

{\bf Keywords:} computational biology, evolution, phylogenetics, BEAST2, 
pirouette, R

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The development of new powerful inference tools, 
such as BEAST [\cite{drummond2007beast}], 
MrBayes [\cite{huelsenbeck2001mrbayes}]
or RevBayes [\cite{hohna2016revbayes}], 
allows to build phylogenetic trees 
from character data (usually nucleotide sequences) extracted 
from extant (but also extinct) organisms.
This has constituted an important step forward 
in our understanding of how species evolve.
Such tools have been increasingly exploited to hypothesize 
and test what the main drivers and modes for diversification are.

Because BEAST performs a Bayesian analysis, 
it does not only need character data but also tree priors, 
describing how the diversification occurs, to return posterior phylogenies.
BEAST uses standard tree priors such as Yule or 
constant-rate birth-death [\cite{nee1994reconstructed}] models.
These models are the most commonly picked in biological research,
as these embody a reasonable level of biological complexity, 
yet are computationally fast enough.
The recent development of BEAST2 [\cite{bouckaert2014beast}] provides, 
unlike its predecessor, the possibility for third-party users 
to use their own new or preferred diversification model, 
if an algorithm is provided to calculate the probability 
of a tree given such a diversification model.
There have been plenty of these models (and their associated probability 
algorithms) 
developed, among others, those in which diversification is 
time-dependent [\cite{nee1994reconstructed}, \cite{rabosky2008explosive}], 
or diversity-dependent [\cite{etienne2011diversity}],
or that can shift [\cite{etienne2012conceptual}, 
\cite{rabosky2014automatic}, \cite{alfaro2009nine}, \cite{laudanno2018sls}].
Other birth-death models treat speciation as a process that takes 
time [\cite{rosindell2010protracted}][\cite{etienne2012prolonging}], 
or that happens in bursts of simultaneous 
events [\cite{laudanno2018mbd}], or that
depends on a trait that has two [\cite{maddison2007estimating}], 
or more [\cite{fitzjohn2012diversitree}] states,
or even allows concealed states [\cite{beaulieu2016detecting}] 
or a combination of all these [\cite{herrera2018detecting}].
Not all of these diversification models are available in BEAST2, 
but the option is available.

When investigating a diversification process,
starting from a given phylogeny, 
the parameter values of the diversification model 
are commonly estimated using maximum likelihood.
To test the effectiveness of the diversification model,
it is standard practice to simulate new phylogenies
from the (maximum likely) parameters and use maximum
likelihood to again estimate the parameters.
An effective diversification model has an
initial parameter estimate that is similar to the control parameter estimates.

Not all diversification models that pass the procedure described above
are used in Bayesian inference.
In Bayesian inference, the likelihood of a diversification 
model, also called tree prior, is computed millions of time.
Due to this, effective but computationally expensive tree priors 
are unused in Bayesian inference.
Also, there may be no need to add such a tree prior to a Bayesian
inference framework: the more informative the data, 
the lower the influence of the tree prior.
Due to this, existing and easy to compute tree priors are most commonly used.
However, it is unknown if it is valid to assume that tree prior choice is 
of low importance.

Here we introduce a method to test the importance of a tree prior.
The method starts with a phylogeny generated by the process of 
a (usually novel) tree prior.
Using existing tree priors, a Bayesian posterior with phylogenies is inferred.
The more simular the posterior phylogenies 
and the original phylogeny are,
the less important the novel tree prior has been.
Only tree priors that are measured to be influential
will be worth the effort and computational burden 
to use in a Bayesian framework.

This method is programmed as an R package called \verb;pirouette;.
\verb;pirouette; is built on \verb;babette; [\cite{bilderbeek2018babette}], 
which calls BEAST2 [\cite{bouckaert2014beast}]. 
With \verb;pirouette;, one has a method to quantify the importance 
of a novel diversification model.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Description}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\verb;pirouette; is written in the R programming language (\cite{R}).
The goal of \verb;pirouette; is to quantify the importance of a tree
prior. It does so by measuring the inference error BEAST2
makes for a given reconstructed phylogeny, 
simulated under the mechanisms of a (usually novel) diversification model, 
which we will refer to as the 'generative tree prior', $\mathit{P}$. 
Such a study will typically construct phylogenies 
for different combinations of the diversification model's parameters, 
to assess under which scenarios the error made by BEAST2 cannot be neglected. 
Although a well-conducted theoretical study uses many replicates, 
this article only shows examples with one replicate, for the sake of clarity.

\verb;pirouette; is very flexible and allows the user 
to specify a big variety of custom settings. 
These settings can be grouped in macro-sections, 
according to how they operate in the pipeline. 
We summarize them in tables~\ref{tab:options} and~\ref{tab:definitions}.

Although many possible tests can be performed, 
we present here what we think are the five most important 
example research questions (ERQ) that \verb;pirouette; can answer. 
For these ERQs we will quantify the importance of a tree prior.

\begin{sidewaystable}
\centering
  \begin{tabular}{|@{}c|p{2.5cm}|p{9cm}|p{4.5cm}@{}|}
    \hline
    \centering
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \textbf{Symbol} & \textbf{Sub-argument} & \textbf{Description} & 
\textbf{Possible values} \\ 
    \hline
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{t_{p}}$ & tree prior & Macroevolutionary speciation model 
      & Yule, BD, CBS, CCP, CEP \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{c_{m}}$ & clock model & Clock for the DNA mutation rates 
      & strict, relaxed log-normal \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{s_{m}}$ & site model & Nucleotide substitution model 
      & JC, HKY, TN93, GTR \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{m_{r}}$ & mutation rate & Pace at which mutation occurs 
      & $m_{r} \in \mathbb{R}_{>0}$\\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{r_{s}}$ & root sequence & DNA sequence at the root of the tree 
      & any combination of a, c, g, t \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{m_{p}}$ & MRCA prior & Prior knowledge of a most recent common 
      ancestor & see documentation \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{m_{s}}$ & MCMC settings & Markov Chain Monte Carlo algorithm setup 
      & see documentation \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{m_{t}}$ & model type & Criterion to select an inference model 
      & Generative, Candidate \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{r_{i}}$ & run if & Condition under which an inference model is 
      used & Always, Best candidate\\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{e_{f}}$ & error function & Specifies how to measure the error 
      & nLTT, $|\gamma|$ \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{t_{m}}$ & twin tree prior & Sets the tree prior for the twin tree 
      & Yule, BD \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \hline
  \end{tabular}
  \caption{
    Most important parameter options. 
    BD = birth death (\cite{nee1994reconstructed}), 
    CBS = coalescent Bayesian Skyline (\cite{drummond2005bayesian}), 
    CCP = coalescent constant-population, 
    CEP = coalescent exponential-population,
    JC = Jukes-Cantor 1969 (\cite{jukes1969evolution}), 
    HKY = Hasegawa, Kishino and  Yano (\cite{hasegawa1985dating}), 
    TN93 = \cite{tamura1993estimation}, 
    GTR = Generalised time-reversible model (\cite{tavare1986some}).
  }
  \label{tab:options}
\bigskip

  \begin{tabular}{|@{}c|p{4cm}|p{9cm}|p{3cm}@{}|}
    \hline
    \centering
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \textbf{Symbol} & \textbf{Macro-argument} & \textbf{Description} 
      & \textbf{Includes} \\
    \hline
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{P}$ & Generative tree prior & Diversification model to generate 
      the 'true' tree & $\mathit{t_{p}}$ \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{A}$ & Alignment model & Model used to simulate the 
      alignment ($\mathit{c_{m}}$ must be strict) 
      & $\mathit{c_{m}}, \mathit{s_{m}}, \mathit{m_{r}}, \mathit{r_{s}}$ \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{G}$ & Generative model & The full setting to produce BEAST2 input 
      data & $\mathit{P}$, $\mathit{A}$ \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{I}$ & Inference model & Phylogenetic inference model to run BEAST2 
      & $\mathit{c_{m}}, \mathit{s_{m}}, \mathit{t_{p}}, 
      \mathit{m_{p}}, \mathit{m_{s}}$ \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{C}$ & Inference conditions & Conditions under which $\mathit{I}$ 
      is used in the inference & $\mathit{m_{t}}, \mathit{r_{i}}$\\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{X}$ & Experiment & Full setting for the BEAST2 run & 
      $\mathit{I},\mathit{C}$ \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{E}$ & Error measure parameters & Errors measurement setup & 
      $\mathit{e_{f}}$\\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{T}$ & Twinning parameters & Twin tree creation procedure & 
      $\mathit{t_{m}}$\\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \hline 
  \end{tabular}
  \caption{
    Definitions of terms and relative symbols used in the main text and in 
    Fig~\ref{fig:pipeline}.
  }
  \label{tab:definitions}
  \giovanni{
    Should we rename "Twinning Parameters" to "Twinning model"?
  }
  \richel{
    As the twinning parameters contain more than only the twin model's tree 
    prior, I suggest to keep it 'twinning parameters'
  }
  \giovanni{
    Should we rename "twin model" to "twin prior"?
  }
  \richel{
    I think that's a great idea! I would prefer the name 'twin tree prior' even
    better
  }
\end{sidewaystable}

\subsection{pirouette's pipeline}

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{workflow.png}
  \caption{
    \texttt{pirouette} pipeline. 
    The pipeline starts from a phylogeny (1a) simulated by generative prior 
$\mathit{P}$.
    The phylogeny is converted to an alignment (2a) using alignment model 
$\mathit{A}$. 
    The user supplies one or more experiments.
    For each experiment $\mathit{X}$ 
    (a tuple of inference model $\mathit{I}$ and condition $\mathit{C}$),
    if its condition $\mathit{C}$ is 
    satisfied (which can depend on the alignment), 
    its inference model $\mathit{I}$ is used in the next step (3a).
    The allowed inference models use the alignment (2a) 
    to create a BEAST2 posterior of (parameter estimates and) phylogenies (4a). 
    Each of the posteriors' trees is compared to the true phylogeny (1a) 
    using error measure $\mathit{E}$, 
    resulting in an error distribution (5a). 
    Optionally, a twin phylogeny (1b) can be generated from the original 
    phylogeny (1a) using twinning parameters $T$, where $T$ may depend on
    the experiments \richel{improve}.
    From a twin phylogeny, the same intermediate steps are taken,
    resulting in a twin error distribution (5b)
  }
  \label{fig:pipeline}
\end{figure}

The pipeline is summarized by the following steps:
\begin{enumerate}
    \item for a given phylogeny an alignment is simulated under a known 
alignment model;
    \item for this alignment, according to the specified inference conditions, 
an inference model is picked (which may be different from the generative 
alignment model);
    \item the alignment is then used as BEAST2 input to infer a posterior 
distribution of phylogenies;
    \item posterior phylogenies are compared with the given original phylogeny 
to estimate the error made, according to the error measure specified by the 
user;
\end{enumerate}
The pipeline is visualized in Fig.~\ref{fig:pipeline}. 
There is also the option to generate a 'twin tree', 
that goes through the same pipeline. 
The utility of this twin tree will be explained below.

The first step simulates a DNA alignment from a given 
phylogeny (Fig.~\ref{fig:pipeline}, 1a $\rightarrow$ 2a).
This operation is performed according to the alignment model 
specified by the user, which consists in a root DNA sequence, 
a DNA mutation rate and a site model.
A strict clock model is used as the only clock model
type, as the currently used alignment simulation algorithm
only supports a mutation rate that is equal across all the branches.
This step is relatively fast, 
but longer DNA alignments will noticeably slow down the inference step.

The second step (Fig.~\ref{fig:pipeline}, 2a $\rightarrow$ 3a)
starts from experiments as set up by the user. 
We define an experiment as the combination of an inference model 
and the conditions to actually use it in the inference step.
The inference model can be selected in two ways, either using the
generative model or selecting one or more candidates out of a list of candidate 
inference models.
An experiment can use the generative model as the inference model, if the 
generative model is known.
An experiment uses a candidate model if either the generative model is unknown,
or there is an interest in seeing how well non-generative models compete with
the generative model.
Typically, a generative model and/or a best candidate model are used in the 
actual inference.
When using an experiment with a generative model,
it must use the same site and clock model used in the alignment simulation,
as well as the tree prior that underlied the creation of the given phylogeny. 
When using experiments with candidate models,
\verb;pirouette; selects the best candidate based
on the evidence (i.e. marginal likelihood) for the model given the alignment.
The evidence of an inference model is estimated using a nested sampling (NS)
approach, as described in \cite{maturana2018model}. The nested sampling is
performed by \verb;mcbette; [\cite{mcbette}], that calls the 'NS' BEAST2 
package. 
Using BEAST2 packages (in a scripted way) can only be done under Linux and Mac,
therefore the selection of candidate models is not available directly on 
Windows. 
However, even on Windows systems there is the option to perform the model 
selection from browser, using \verb;mcbette;, yet the computation for this 
model selection is
restricted to one hour for free users of the host (Travis CI) continuous
integration service.

The third step infers the actual Bayesian posteriors from the simulated 
alignment (Fig.~\ref{fig:pipeline}, 2a $\rightarrow$ 3a $\rightarrow$ 4a),
using the inference models from the experiments selected in the previous step. 
For each selected experiment, one BEAST2 posterior is inferred, using the 
\verb;babette; [\cite{bilderbeek2018babette}] R package.

The fourth step compares the true tree to each posterior tree, resulting in an 
error distribution (Fig.~\ref{fig:pipeline}, 4a $\rightarrow$ 5a).
The nLTT statistic (\cite{janzen2015approximate}) is used by default, but also 
a user-defined error statistic can be used. 
As an example, \verb;pirouette; supplies one custom error statistic,
that uses an absolute difference in the gamma statistic 
[\cite{pybus2000testing}].
Additionally, the user can specify the proportion of posterior phylogenies to 
discard (i.e. the burn-in), throwing away the first $10\%$
of all phylogenies by default. This burn-in is used to discard
the part of the MCMC chain that has not yet converged to a
representative part of the state space.

\subsection{Twin tree}\label{subsec:twinning}

An optional step is to generate a 'twin tree' $\tau_{I}$
(Fig.~\ref{fig:pipeline}, 1a $\rightarrow$ 1b),
that will be analyzed in the same way as the true tree, following a parallel 
pipeline. We want to produce such tree 'converting' the original tree 
$\tau_{\mathit{P}}$, created according to the generative prior $\mathit{P}$, to 
be in accordance with the chosen inference model $\mathit{I}$:
\begin{align}
    \tau_{\mathit{P}} = (\Vec{t}_{\mathit{P}}, \psi_{\mathit{P}}) 
\xrightarrow[]{\mathit{T}} \tau_{\mathit{I}} = (\Vec{t}_{\mathit{I}}, 
\psi_{\mathit{I}}),
\end{align}
where $\Vec{t}$ are the tree branching times and $\psi$ is tree topology, which 
provides information on where new lineages branch out at each branching time.

To execute the twinning process we exploit the likelihood function 
$L_{\mathit{I}}$ associated with the chosen inference model $\mathit{I}$. We 
find the parameters $\theta^{*}_{\mathit{I}}$ (e.g. speciation and extinction 
rates, in case of a birth-death model) that maximize $L_{\mathit{I}}$ applied 
to the true tree, conditioned on its number of tips $n_{\mathit{P}}$,
\begin{align}
    max[L_{\mathit{I}}(\theta_{\mathit{I}}|\tau_{\mathit{P}}, n_{\mathit{P}})] 
\rightarrow \theta^{*}_{\mathit{I}}.
\end{align}
We use $\theta^{*}_{\mathit{I}}$ to simulate an equal amount $n_{\mathit{I}} = 
n_{\mathit{P}}$ of branching times $\Vec{t}_{\mathit{I}}$ for the twin tree 
$\tau_{\mathit{I}}$, under the process $\mathit{I}$. We then impose that 
$\tau_{\mathit{P}}$ and $\tau_{\mathit{I}}$ must respect the same topology
\begin{align}
    \psi_{\mathit{I}} = \psi_{\mathit{P}}.
\end{align}
Using a twin tree serves as a control: both $\tau_{\mathit{P}}$ and 
$\tau_{\mathit{I}}$ will produce an error distribution. We assume that these 
errors are a natural product of the input tree and the intermediate steps of 
the pipeline. All the factors that can cause such errors are: the branching 
times $\vec{t}$ of the tree (as a consequence of the tree prior), its topology 
$\psi$, the stochasticity induced by simulating the alignment (step 1a 
$\rightarrow$ 2a in Fig.~\ref{fig:pipeline}) and the stochasticity induced by 
the Bayesian inference (step 2a $\rightarrow$ 4a in Fig.~\ref{fig:pipeline}). 
By imposing the same topology and replicating the pipelines a sufficient number 
of times we can sort out all the possible causes that are not a direct 
consequence of the choice of the tree prior.

Potential differences in shape of error distributions will therefore be 
consequences of different tree priors for the two parallel pipelines: the use 
of prior $\mathit{P}$ for the original pipeline and of prior $\mathit{I}$ for 
the twin pipeline. In the second case, though, such error distribution is not 
influenced by the mismatch between how the tree is generated and how it is 
analyzed by BEAST2.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Installation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\verb;pirouette; can be installed easily from CRAN:
\begin{lstlisting}[language=R, floatplacement=ht, frame=single]
install.packages("pirouette")
\end{lstlisting}

For the most up-to-date version, 
one can download and install the package from \verb;pirouette;'s GitHub 
repository:

\begin{lstlisting}[language=R, floatplacement=ht, frame=single]
usethis::install_github("richelbilderbeek/pirouette")
\end{lstlisting}

To start using \verb;pirouette;, load its functions in the global namespace 
first:

\begin{lstlisting}[language=R, floatplacement=ht, frame=single]
library(pirouette)
\end{lstlisting}
Because \verb;pirouette; calls BEAST2, BEAST2 must be installed. 
This can be done from within R, using:

\begin{lstlisting}[language=R, floatplacement=ht, frame=single]
install_beast2()
\end{lstlisting}
For the option to select a best candidate model,
\verb;pirouette; needs the "NS" BEAST2 package [\cite{maturana2018model}].
It can be installed from within R, using:

\begin{lstlisting}[language=R, floatplacement=ht, frame=single]
install_beast2_pkg("NS")
\end{lstlisting}

An overview of \verb;pirouette;'s main functions is shown in 
table~\ref{tab:functions}. 
These functions are demonstrated when in the example code below.
All \verb;pirouette;'s functions are documented,
have a useful example, and have sensible defaults.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}[h]
\centering
\begin{tabular}{ | l | l | }
\hline
\textbf{Name} & \textbf{Description} \\
\hline
\verb;pir_run; & Run \verb;pirouette; \\
\verb;create_pir_params; & Create the \verb;pirouette; parameters \\
\hline
\verb;create_alignment_params; & Create the alignment parameters \\
\verb;create_twinning_params; & Create the twinning parameters \\
\verb;create_experiment; & Create one experiment \\
\verb;create_error_measure_params; & Create the error measurement parameters \\
\hline
\end{tabular}
\caption{pirouette's main functions.}
\label{tab:functions}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Usage: Example Research Question 1}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Question}

\verb;pirouette; quantifies the influence of a new tree prior on BEAST2's 
inference,
by measuring the difference between a given tree and 
phylogenies in a posterior. Due to stochasticity, all posterior trees
will differ from the given phylogeny, even when the generative tree
prior $\mathit{P}$ and alignment model $\mathit{A}$ are the same as 
those in the inference model $\mathit{I}$.
Measuring this difference allows us to know the baseline error
of the \verb;pirouette; pipeline. We therefore define as 'standard tree priors' 
all the tree priors that are available to be used within an inference model 
(see Table~\ref{tab:options}).

Now we can formulate the first example research question that \verb;pirouette; 
can answer:

"What is the inference error made on phylogenies
created by a standard diversification model?"

\subsection{Answer}

In this example we use a standard generative tree prior $\mathit{P_{0}}$, 
namely the Yule (pure-birth) tree prior. 
We choose to use a small tree with six taxa to keep
the calculations short and the figure more readable.
We pick a crown age of ten time units. This value is 
completely arbitrary, but does tie in with the mutation rate 
used in simulating an alignment in the next step.

\begin{lstlisting}[
    language=R, floatplacement=ht, frame=single, 
    label = {lst:create_yule_tree}, 
    caption = {
      Create a Yule tree. 
      The resulting tree is shown in figure~\ref{fig:yule_tree}.
    }
  ]
phylogeny <- create_yule_tree(n_taxa = 6, crown_age = 10)
\end{lstlisting}

\begin{figure}[ht]
  \includegraphics[width=\textwidth]{tree_yule.png}
  \caption{The Yule tree, as created by listing~\ref{lst:create_yule_tree}.}
  \label{fig:yule_tree}
  \giovanni{Remove the text in top-left corner.}
  \richel{Agreed, iff the text matches the code}
  \giovanni{I'd use a sexier topology, purely for esthetical reasons.}
  \richel{Agreed, will pick a sexy RNG seed after create\_yule\_tree is there}
\end{figure}

The first step in \verb;pirouette; is to simulate a DNA alignment from the 
given phylogeny. 
To do so, we must specify an alignment model $\mathit{A}$ which consists of: a 
DNA root sequence $\mathit{r_{s}}$, the DNA mutation rate $\mathit{m_{r}}$, a
site model $\mathit{s_{m}}$ (in this context: a nucleotide substitution model) 
and a clock model $\mathit{c_{m}}$. 
In this example, the DNA root sequence consists in four blocks of 250 
nucleotides each, 
where the per-nucleotide mutation rate is 0.1 mutations per unit time.
We use a Jukes-Cantor (JC, \cite{jukes1969evolution}) site model
and a strict clock model as these are the simplest.
A Jukes-Cantor site model
assumes that mutation rates between nucleotides are equal and constant. 
A strict clock model assumes that the mutation rates 
of all species are equal and constant.

\begin{lstlisting}[
    language=R,
    floatplacement=ht, frame=single,
    label = {lst:create_alignment}, 
    caption = {Create an alignment.}
  ]
alignment_params <- create_alignment_params(
  root_sequence = create_blocked_dna(length = 1000),
  mutation_rate = 0.1,
  site_model = create_jc69_site_model(),
  clock_model = create_strict_clock_model()
)
\end{lstlisting}

As the site and clock models used here are also the defaults, 
these function arguments can be safely omitted: we just show these
explicitly for clarity.

In the second step we state our experiment.
We define an experiment $\mathit{X}$ as a combination of an inference model 
$\mathit{I}$
and conditions $\mathit{C}$.
In this example we pick $\mathit{I}$ to be the same inference model as the 
generative one,
which is the chosen Yule tree prior $\mathit{P_{0}}$, as well as site and clock 
models defined in $\mathit{A}$, respectively JC and strict.
We specify in $\mathit{C}$ that the experiment will always be run.

\begin{table}
  \begin{tabular}{ | c | c | c | l | }
    \hline
    \textbf{model type} & \textbf{run if} & \textbf{measure evidence} & 
\textbf{inference model} \\ 
    \hline
    generative & always & no & JC, strict, Yule \\
    \hline
  \end{tabular}
  \caption{
    Inference conditions and inference model for ERQ 1.
    JC: Jukes-Cantor site model.
    strict: strict clock model.
    Yule: Yule (pure-birth) tree prior.
  }
  \label{tab:RQ1}
\end{table}

Listing~\ref{lst:create_generative_experiment_explicit} shows how to
set up this experiment:

\begin{lstlisting}[
  language=R, 
  floatplacement=ht, frame=single,
  label = {lst:create_generative_experiment_explicit},
  caption = {
    Create the experiments to answer research question 1, 
    using explicit arguments
  }
]
experiment <- create_experiment(
  inference_conditions = create_inference_conditions(
    model_type = "generative", 
    run_if = "always"
  ), 
  inference_model = create_inference_model(
   site_model = create_jc69_site_model(), 
   clock_model = create_strict_clock_model(), 
   tree_prior = create_yule_tree_prior()
  )
)
\end{lstlisting}

Experiments must be bundled in a list to work, even if only one is provided, as 
in this case:

\begin{lstlisting}[
  language=R, 
  floatplacement=ht, frame=single,
  label = {lst:create_generative_experiment},
  caption = {
    Create the experiments to answer research question 1.
  }
]
experiments <- list(experiment)
\end{lstlisting}

The objects we just created need to be bundled
in a \verb;pirouette; parameters structure,
together with error measurement parameters $\mathit{E}$ and
twinning parameters $\mathit{T}$.

We use the default error measurement
parameters, which uses a burn-in of 10\% and the nLTT statistic to
measure the difference between phylogenies. For clarity,
we create this setup explicitly here:

\begin{lstlisting}[
  language=R, floatplacement=ht, frame=single,
  label = {lst:create_error_measure_params},
  caption = Calling \texttt{create\_error\_measure\_params}
]
error_measure_params <- create_error_measure_params(
  burn_in_fraction = 0.1, 
  error_function = get_nltt_error_function()
)
\end{lstlisting}

We will not use twinning
here yet.  
All these arguments are bundled
and checked by \verb;create_pir_params;:

\begin{lstlisting}[
  language=R, floatplacement=ht, frame=single,
  label = {lst:create_pir_params},
  caption = Calling \texttt{create\_pir\_params}
]
pir_params <- create_pir_params(
  alignment_params = alignment_params,
  experiments = experiments,
  error_measure_params = error_measure_params,
  twinning_params = NA
)
\end{lstlisting}

Now we can use our Yule tree and \verb;pir_params; to measure 
the inference error made on phylogenies
created by a standard diversification model:

\begin{lstlisting}[
  language=R, floatplacement=ht, frame=single,
  label = {lst:pir_run},
  caption = Calling \texttt{pir\_run}
]
errors <- pir_run(
  phylogeny = phylogeny,
  pir_params = pir_params
)
\end{lstlisting}

The error distribution returned by \verb;pir_run; is stored as a data frame. 
\richel{
  I suggest \texttt{pir\_run} to return a list of run experiments,
  as described in Issue 225, 
  \url{https://github.com/richelbilderbeek/pirouette/issues/225}
}
It can be directly plotted using \verb;pir_plot;:

\begin{lstlisting}[
  language=R, floatplacement=ht, frame=single,
  label = {lst:pir_plot},
  caption = Calling \texttt{pir\_plot}
]
pir_plot(errors)
\end{lstlisting}

\begin{figure}[ht]
  \includegraphics[width=\textwidth]{example_1_errors.png}
  \caption{
    Example 1: the error BEAST2 makes from a phylogeny 
    when the generative and inference model are the same.
  }
  \label{fig:example_1}
\end{figure}

The error distribution obtained is shown as a violin plot
in figure~\ref{fig:example_1}
\richel{
  I suggest to use a histogram (or density plot) instead:
  I like how the error distribution in figure 1 looks.
}
.

This error distribution has little relevance on its own, 
as we constructed the experiment in such a way $\mathit{I}$ 
matches with the generative model given by $\mathit{A}$ and $\mathit{P_{0}}$.
However it will become clear when compared to the errors obtained from a 
phylogeny 
derived from a non-standard diversification process $\mathit{P}$.

By doing so, this error distribution will serve as a control,
as it can be seen as the baseline error caused by stochasticity.
If the two error distributions obtained from $\mathit{P_{0}}$ and $\mathit{P}$ 
will appear similar, the error induced by the choice of a non-standard 
speciation process will be of little influence.
Given the necessity of such comparison,
we made it an inherent part of \verb;pirouette;. 
We call it 'twinning' and we use it in ERQ 3.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Usage: Example Research Question 2}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Question}

>>>>

\title{
  Indirect evidence white is actually a funny color
}
\author[1]{Rich\`el J.C. Bilderbeek}
\affil[1]{Groningen Institute for Evolutionary Life Sciences, University of 
Groningen, Groningen, The Netherlands}

@@
